\chapter{Building, Running, and Deploying}
\label{chap:build}

\section{Dependencies at a Glance}

From \verb|Cargo.toml| (the binary is \verb|seal-server|, Rust edition 2021):

\begin{center}
\begin{longtable}{@{}llp{0.42\textwidth}@{}}
\toprule
\textbf{Crate} & \textbf{Version} & \textbf{Role} \\
\midrule
\endhead
\verb|axum| & 0.8 & Web framework (features \verb|ws|, \verb|macros|). \\
\verb|tokio| & 1 & Async runtime (\verb|full|). \\
\verb|tower| / \verb|tower-http| & 0.5 / 0.6 & Middleware; \verb|fs| + \verb|trace|. \\
\verb|serde| / \verb|serde_json| / \verb|serde_yml| & 1 / 1 / 0.0.13 & Serialization (JSON, YAML). \\
\verb|dotenvy| & 0.15 & \verb|.env| loading. \\
\verb|bcrypt| & 0.19 & Password hashing. \\
\verb|jsonwebtoken| & 10 & JWTs (\verb|rust_crypto|). \\
\verb|lancedb| & 0.30 & Embedded database. \\
\verb|arrow-array| & 58 & Arrow arrays (version-locked to \verb|lancedb|). \\
\verb|uuid| & 1 & Ids (\verb|v4|). \\
\verb|regex| & 1 & Validation patterns. \\
\verb|tracing| / \verb|tracing-subscriber| & 0.1 / 0.3 & Logging. \\
\verb|thiserror| / \verb|anyhow| & 2 / 1 & Errors. \\
\verb|include_dir| & 0.7 & Asset bundling. \\
\verb|mime_guess| & 2 & Static-asset content types. \\
\bottomrule
\end{longtable}
\end{center}

Dev-dependencies power the integration suite: \verb|reqwest| (HTTP client, rustls),
\verb|tokio-tungstenite| (WebSocket client), and \verb|tempfile| (throwaway
databases).

\begin{quote}
\textbf{Arrow / LanceDB coupling.} \verb|arrow-array| is pinned to the major
version LanceDB expects (58 for \verb|lancedb| 0.30). Bumping one usually means
bumping the other in lock-step; a mismatch produces confusing trait-resolution
errors at the Arrow boundary.
\end{quote}

\section{The Makefile}

\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Target} & \textbf{Action} \\
\midrule
\verb|make dev| / \verb|make build| & \verb|cargo build| (debug). \\
\verb|make release| & \verb|cargo build --release|. \\
\verb|make server| & \verb|cargo run|. \\
\verb|make test| / \verb|make test-quiet| & Run the test suite. \\
\verb|make bots| & Run the Python bot simulation against a live server. \\
\verb|make clean| & \verb|cargo clean| + remove Tauri and \verb|__pycache__|. \\
\verb|make reset-db| & Delete \verb|data/chat.lance|. \\
\verb|make bundle-sodium| & Rebuild the vendored libsodium bundle. \\
\bottomrule
\end{tabular}
\end{center}

\section{Running Locally}

\begin{lstlisting}
make dev && make server         # http://localhost:8000
# or, self-contained:
cargo run --release
\end{lstlisting}

The release binary at \verb|target/release/seal-server| is self-contained ---
templates, static assets, and the default config are compiled in. To deploy it,
copy the binary, set \verb|JWT_SECRET| and \verb|DATABASE_PATH|, and run.

\section{Docker}

The \verb|Dockerfile| is a two-stage build:

\begin{enumerate}
  \item \textbf{Builder} (\verb|rust:1-bookworm|). Installs the native toolchain
        LanceDB needs --- \verb|protobuf-compiler|, \verb|libprotobuf-dev|,
        \verb|cmake|, \verb|pkg-config|, \verb|libssl-dev| --- and sets
        \verb|PROTOC| / \verb|PROTOC_INCLUDE| because \verb|lance-encoding|'s
        build script invokes \verb|protoc| and imports the well-known
        \verb|google/protobuf/empty.proto|. Then \verb|cargo build --release|.
  \item \textbf{Runtime} (\verb|debian:bookworm-slim|). Copies just the binary,
        adds \verb|ca-certificates|, creates \verb|/app/data|, and sets
        \verb|APP_HOST=0.0.0.0|, \verb|APP_PORT=8080|,
        \verb|DATABASE_PATH=/app/data/chat.lance|. Exposes 8080.
\end{enumerate}

\section{Continuous Integration}

Two GitHub Actions workflows live in \verb|.github/workflows/|:

\begin{itemize}
  \item \textbf{\texttt{release.yml}} --- on every push to \verb|main| (and
        manually), cross-compiles the server for Linux x86\_64
        (\verb|x86_64-unknown-linux-gnu|) and macOS Apple Silicon
        (\verb|aarch64-apple-darwin|), packages each as a \verb|.tar.gz|, and
        publishes them as a GitHub \emph{pre-release} tagged
        \verb|build-YYYYMMDD-<short-sha>|. Installs \verb|protoc| and caches the
        Cargo build.
  \item \textbf{\texttt{coverage.yml}} --- on pushes to \verb|main|, PRs, and
        manual runs, uses \verb|cargo llvm-cov| (with
        \verb|llvm-tools-preview|) to produce \verb|lcov.info|, an HTML report,
        and a text summary published to the workflow run.
\end{itemize}

\section{Deploying to DigitalOcean}

\verb|scripts/deploy-digitalocean.sh| is an interactive helper for the App
Platform. It prompts for the app name, GitHub repo/branch, region, and instance
size, generates an app-spec YAML, and runs \verb|doctl apps create --wait| (or
\verb|update|), optionally enabling \verb|deploy_on_push|. Be aware that the App
Platform filesystem is \emph{ephemeral} by default: a local
\verb|DATABASE_PATH| is wiped on redeploy. For durable data, attach a persistent
volume --- or use the object-storage backend
(Chapter~\ref{chap:object-storage}) so LanceDB persists to a bucket.

\section{Auxiliary Scripts and the Desktop Client}

\begin{itemize}
  \item \verb|scripts/bots.py| --- a deterministic simulation (idempotent
        accounts and channels, randomized messages) that drives the server with
        \emph{real} PyNaCl X25519 encryption, proving cross-implementation
        compatibility with the JS client. Run via \verb|make bots| (which uses
        \verb|uv|).
  \item \verb|scripts/bundle-sodium.sh| --- rebuilds the vendored libsodium
        bundle shipped in \verb|static/|.
  \item \verb|desktop/| --- a Tauri v2 native client (Rust + system WebView)
        wrapping the same web front end.
\end{itemize}
